---
title: "lab10"
author: "Nicole Tang"
format: html
embed-resources: true
fig-width: 6
fig-height: 4
---


# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
library(RSQLite)
library(DBI)

# Initialize a temporary in memory database
con <- dbConnect(SQLite(), ":memory:")
```



```{r}
# install.packages(c("RSQLite", "DBI"))

library(RSQLite)
library(DBI)

# Initialize a temporary in memory database
con <- dbConnect(SQLite(), ":memory:")

# Download tables
actor <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/actor.csv")
rental <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/rental.csv")
customer <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/customer.csv")
payment <- read.csv("https://raw.githubusercontent.com/ivanceras/sakila/master/csv-sakila-db/payment_p2007_01.csv")

# Copy data.frames to database
dbWriteTable(con, "actor", actor)
dbWriteTable(con, "rental", rental)
dbWriteTable(con, "customer", customer)
dbWriteTable(con, "payment", payment)
```

```{r}
dbListTables(con)
```

TIP: You can use the following QUERY to see the structure of a table

```{sql, connection=con}
PRAGMA table_info(actor)
```

SQL references:

https://www.w3schools.com/sql/

# Exercise 1

Edit the code below to retrieve the actor ID, first name and last name for all actors using the `actor` table. Sort by last name and then by first name (note that the code chunk below is set up to run SQL code rather than R code).

```{sql, connection=con}
SELECT actor_id, first_name, last_name
FROM actor
ORDER by last_name asc, first_name asc
```

# Exercise 2

Retrieve the actor ID, first name, and last name for actors whose last name equals 'WILLIAMS' or 'DAVIS'.

```{sql, connection=con}
SELECT actor_id, first_name, last_name
FROM actor
WHERE last_name IN ('WILLIAMS', 'DAVIS')
```

# Exercise 3

Write a query against the `rental` table that returns the IDs of the customers who rented a film on July 5, 2005 (use the rental.rental_date column, and you can use the `date()` function to ignore the time component). Include a single row for each distinct customer ID. 

```{sql, connection=con}
SELECT DISTINCT rental_id
FROM rental as r
WHERE date(r.rental_date) = '2005-07-05'
```
# Exercise 4

## Exercise 4.1

Construct a query that retrieves all rows from the `payment` table where the amount is either 1.99, 7.99, 9.99.

```{sql, connection=con}
SELECT *
FROM payment
WHERE amount IN (1.99, 7.99, 9.99)
```

## Exercise 4.2

Construct a query that retrieves all rows from the `payment` table where the amount is greater then 5.

```{sql, connection=con}
SELECT *
FROM payment
WHERE amount > 5.0
```

## Exercise 4.2

Construct a query that retrieves all rows from the `payment` table where the amount is greater then 5 and less then 8.

```{sql, connection=con}
SELECT *
FROM payment
WHERE amount > 5.0 AND amount < 8.0
```
# Exercise 5

Retrieve all the payment IDs and their amounts from the customers whose last name is 'DAVIS'.

```{sql, connection=con}
SELECT p.payment_id, p.amount
FROM payment as p
  INNER JOIN customer as c on c.customer_id = p.customer_id
WHERE c.last_name = "DAVIS"
```

# Exercise 6

## Exercise 6.1

Use `COUNT(*)` to count the number of rows in `rental`.

```{sql, connection=con}
SELECT count(*) as count
FROM rental
```

## Exercise 6.2

Use `COUNT(*)` and `GROUP BY` to count the number of rentals for each `customer_id`.

```{sql, connection=con}
SELECT customer_id, count(*) as count
FROM rental
GROUP BY customer_id
```
## Exercise 6.3

Repeat the previous query and sort by the count in descending order.

```{sql, connection=con}
SELECT customer_id, count(*) as count
FROM rental
GROUP BY customer_id
ORDER BY count desc
```

## Exercise 6.4

Repeat the previous query but use `HAVING` to only keep the groups with 40 or more.

```{sql, connection=con}
SELECT customer_id, count(*) as count
FROM rental
GROUP BY customer_id
HAVING count >= 40
ORDER BY count desc
```

# Exercise 7

Write a query that calculates a number of summary statistics for the payment table using `MAX`, `MIN`, `AVG` and `SUM`

```{sql, connection=con}
SELECT MAX(amount) AS max, MIN(amount) AS min, AVG(amount) AS avg, SUM(amount) AS total
FROM payment
```

## Exercise 7.1

Modify the above query to do those calculations for each `customer_id`.

```{sql, connection=con}
SELECT c.customer_id, MAX(p.amount) AS max, MIN(p.amount) AS min, AVG(p.amount) AS avg, SUM(p.amount) AS total
FROM payment as p
  INNER JOIN customer as c on c.customer_id = p.customer_id
GROUP BY c.customer_id
```
## Exercise 7.2

Modify the above query to only keep the `customer_id`s that have more then 5 payments.

```{sql, connection=con}
SELECT c.customer_id, count(c.customer_id) as purchases, MAX(p.amount) AS max, MIN(p.amount) AS min, AVG(p.amount) AS avg, SUM(p.amount) AS total
FROM payment as p
  INNER JOIN customer as c on c.customer_id = p.customer_id
GROUP BY c.customer_id
HAVING purchases > 5
```
# Cleanup

Run the following chunk to disconnect from the connection.

```{r}
# clean up
dbDisconnect(con)
```