---
title: "Assignment_1"
author: "Nicole Tang"
format: html
embed-resources: true
fig-width: 6
fig-height: 4
---
```{r setup, message=FALSE, echo=FALSE, warning=FALSE}
library(data.table)
library(leaflet)
library(tidyverse)
library(data.table)
library(R.utils)
library(dplyr)
library(ggplot2)
library(lubridate)
library(gganimate)
library(ggforce)
```

# Due Date

This assignment is due by 11:59pm Pacific Time, September 27th, 2024.


# Assignment Description

We will work with air pollution data from the U.S. Environmental Protection Agency (EPA). The EPA has a national monitoring network of air pollution sites that The primary question you will answer is whether daily concentrations of PM$_{2.5}$ (particulate matter air pollution with aerodynamic diameter less than 2.5 $\mu$m) have decreased in California over the last 20 years (from 2002 to 2022).

A primer on particulate matter air pollution can be found [here](https://www.epa.gov/pm-pollution/particulate-matter-pm-basics).

Your assignment should be completed in Quarto or R Markdown.

# Steps

1. Given the formulated question from the assignment description, you will now conduct EDA Checklist items 2-4. First, download 2002 and 2022 data for all sites in California from the [EPA Air Quality Data website](https://www.epa.gov/outdoor-air-quality-data/download-daily-data). Read in the data using `data.table()`. For each of the two datasets, check the dimensions, headers, footers, variable names and variable types. Check for any data issues, particularly in the key variable we are analyzing. Make sure you write up a summary of all of your findings.

```{r}
#read in data
air2002 <- fread("ad_viz_plotval_data_2022.csv")
air2022 <- fread("ad_viz_plotval_data_2002.csv")

```
```{r}
#observe data
dim(air2002)
head(air2002)
tail(air2002)
colnames(air2002)
str(air2002)
summary(air2002)
colSums(is.na(air2002))
```
**Dimensions: 22x59756**
**Column Names(type): Date(chr), Source(chr), Site ID(int), POC(int), Daily Mean PM2.5 Concentration(num), Units(chr), Daily AQI Value(int), Local Site Name(chr), Daily Obs Count(int), Percent Complete(num), AQS Parameter Code(int), AQS Parameter Description (chr), Method Code(int), Method Description(chr), CBSA Code(int), CBSA Name(chr), State FIPS Code(int), State(chr), County FIPS Code(int), County(chr), Site Latitude(num), Site Longitude(num)**
**No NAs in the data set**

```{r}
dim(air2022)
head(air2022)
colnames(air2022)
tail(air2022)
str(air2022)
summary(air2022)
colSums(is.na(air2022))
```
**Dimensions: 22x15976**
**Column Names(type): Date(chr), Source(chr), Site ID(int), POC(int), Daily Mean PM2.5 Concentration(num), Units(chr), Daily AQI Value(int), Local Site Name(chr), Daily Obs Count(int), Percent Complete(num), AQS Parameter Code(int), AQS Parameter Description (chr), Method Code(int), Method Description(chr), CBSA Code(int), CBSA Name(chr), State FIPS Code(int), State(chr), County FIPS Code(int), County(chr), Site Latitude(num), Site Longitude(num)**
**No NAs in the data set**

2. Combine the two years of data into one data frame. Use the Date variable to create a new column for year, which will serve as an identifier. Change the names of the key variables so that they are easier to refer to in your code.
```{r}
air <- rbind(air2002, air2022)

summary(air$Date)

air <- air %>%
  mutate(Year = year(as.Date(Date, format = "%m/%d/%Y")))

air <- air %>%
  rename(PM2.5 = "Daily Mean PM2.5 Concentration")
air <- air %>%
  rename(lat = "Site Latitude")
air <- air %>%
  rename(long = "Site Longitude")
air <- air %>%
  rename(site = "Local Site Name")

```


3. Create a basic map in leaflet() that shows the locations of the sites (make sure to use different colors for each year). Summarize the spatial distribution of the monitoring sites.

```{r}
pal <- colorFactor(c("lightgreen", "purple"), domain = unique(air$Year))

# Create a leaflet map
leaflet(data = air) %>%
  addTiles() %>%
  addCircleMarkers(
    ~long, ~lat,  
    color = ~pal(Year),     
    radius = 5,             
    fillOpacity = 0.7 
  ) %>%
  addLegend("bottomright", pal = pal, values = ~Year,
            title = "Year",
            opacity = 1)

```
**The graph shows that there are stations all throughout California that have been relaying data in 2002 and 2022.**
```{r}
sum <- air %>%
  group_by(Year) %>%
  summarize(
    Mean = mean(PM2.5, na.rm = TRUE),   
    Median = median(PM2.5, na.rm = TRUE), 
    Min = min(PM2.5, na.rm = TRUE),       
    Max = max(PM2.5, na.rm = TRUE),       
    Count = n()                            
  )
print(sum)

```
**For 2002: Mean: 16.12, Min:0, Max: 104.3, Count:15976**
**For 2022: Mean: 8.43, Min:-6.7, Max:302.5, Count:59756**
**2002 had a higher mean PM2.5 concentration. 2022 had a higher max PM2.5 value. There are more data points from 2022.**


4. Check for any missing or implausible values of PM$_{2.5}$ in the combined dataset. Explore the proportions of each and provide a summary of any temporal patterns you see in these observations.

```{r}
sum(is.na(air$PM2.5))

ggplot(data = air, aes(x = as.factor(Year), y = PM2.5, fill = as.factor(Year))) + 
  geom_boxplot(outlier.colour = "lightblue") +
  labs(title = "Box Plot of PM2.5 by Year",
       x = "Year",
       y = "PM2.5 Concentration (µg/m³)")

sum <- air %>%
  group_by(Year) %>%
  summarize(
    Mean = mean(PM2.5, na.rm = TRUE),   
    Median = median(PM2.5, na.rm = TRUE), 
    Min = min(PM2.5, na.rm = TRUE),       
    Max = max(PM2.5, na.rm = TRUE)
  )
print(sum)
```
**There appears to be a negative value of PM2.5 in 2022. 2022 also appears to have a very high mx but there are multiple high values so it is possibly valid. The average PM2.5 has also halved since 2002.**

5. Explore the main question of interest at three different spatial levels. Create exploratory plots (e.g. boxplots, histograms, line plots) and summary statistics that best suit each level of data. Be sure to write up explanations of what you observe in these data.

    - state
    - county
    - sites in Los Angeles
    
```{r}
ggplot(air, aes(x = State, y = PM2.5, fill = as.factor(Year))) + 
  geom_boxplot(na.rm = TRUE) + 
  labs(x = "State", 
       y = "PM2.5 Concentration (µg/m³)", 
       title = "Box Plot of PM2.5 by State") + 
  theme_minimal() 
```
**Based on this box plot, you can see that 2022 has higher max values while 2002 has a higher average**
```{r}
ggplot(air, aes(x = State, y = PM2.5, fill = as.factor(Year))) + 
  geom_violin(na.rm = TRUE) + 
  labs(x = "State", 
       y = "PM2.5 Concentration (µg/m³)", 
       title = "Violin Plot of PM2.5 by State") + 
  theme_minimal()
```
**Based on this box plot, you can see that 2022 has higher max values while 2002 has a higher average. You can also see that 2002 is more concentrated around the mean.**

```{r}
ggplot(air, aes(x = County, y = PM2.5, fill = as.factor(Year))) + 
  geom_bar(stat = "summary", fun = mean, position = "dodge", na.rm = TRUE) + 
  labs(x = "County", 
       y = "Average PM2.5 Concentration (µg/m³)", 
       title = "Average PM2.5 by County") + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
**This bar graph shows the difference in PM2.5 by year in the various counties. You can observe that 2002 has higher averages than 2002 in all counties**
    
```{r}
losAngeles <- air %>%
  filter(County == "Los Angeles")

ggplot(losAngeles, aes(x = site, y = PM2.5, color = as.factor(Year))) + 
  stat_summary(fun = mean, geom = "point", size = 3, na.rm = TRUE) +  
  stat_summary(fun = mean, geom = "line", aes(group = Year), na.rm = TRUE) +  
  labs(x = "Site in LA", 
       y = "Average PM2.5 Concentration (µg/m³)", 
       title = "Average PM2.5 by Site in Los Angeles") + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
**This line graph shows the difference in average PM2.5 by year in the sites in LA. 2002 is consistantly has a higher average than 2022.**
    

---

This homework has been adapted from the [case study](https://bookdown.org/rdpeng/exdata/data-analysis-case-study-changes-in-fine-particle-air-pollution-in-the-u-s-.html) in Roger Peng's [Exploratory Data Analysis with R](https://bookdown.org/rdpeng/exdata/)
